The dynamics of risk perception in a Mediterranean agroecosystem

### 2020-02-20 manuscript/manuscript.Rmd
working figure crossref

bibliography: bibliography.bib
bibstyle: spbasic
# bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
header-includes:
- \usepackage{float}
- \floatplacement{figure}{tb}
---

@@

We can treat this problem elegantly using Bayesian statistics using online Bayesian learning from bernoulli observations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. bernoulli distribution) with parameter theta reprenting the risk of drought. Uncertainty in the value of theta can be represented with a beta distribution, constrained between 0 and 1. We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief \@ref(beta-prior). What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely. As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of experience.

```{r beta-prior, fig.width = 5, fig.height = 3, fig.cap = "Development of an individual's perceived drought risk with time, assuming a fixed drought risk of 0.5. Beliefs are represented as Beta probabiity distributions, and the increased certainty with age reflects the varying effective sample size of the Beta prior."}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

@@

# Drought risks in the eastern Mediterranean {#sec:2}

Estimates of drought risk in the eastern Mediterranean during the the past 10,000 years were derived from outputs of the TraCE-21k paleoclimate simulation available on the National Center for Atmospheric Research. TraCE-21K is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model (GCM) to recreate the transient response of the global climate system to changes in the Earth's orbit and greenhouse gasses over the past 22,000 years from the Last Glacial Maximum to the present [@He2011]. It generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of external climate drivers (e.g. orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six hourly timescale, and model outputs are archived at a monthly resolution.

@@

```{r, fig.height = 2, fig.width = 6, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period. The dashed line indicates the level of risk beyond which one would plant barley over wheat to maximize subjective expected yields, given the payoffs defined in Table 1.'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  filter(year != min(year)) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, size = 1) +
  geom_col(width = 38)  +
  scale_fill_distiller(palette = 'PuOr') +
  scale_color_distiller(palette = 'PuOr')+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

The bias-corrected wet season precipitation outputs from TraCE-21k show a relatively stationary distribution over the past 10,000 years, with slightly drier conditions during the early-mid Holocene transition at c.a. 8ka BP (Figure 5). The utility of a diversified wheat-barley crop is immediately apparent; rainfall often fails to meet the 300mm threshold for a productive wheat crop during the entire 10,000 year span, but drops below the threshold for barley only a handful of years.

The risk of crop failure due to drought in any given year varies between 10\% and 20\%, punctuated by one to four century-long events where risk drops as low as 6\% and or exceeds 25\% (Figure 6). Centuries of high drought risk cluster around known periods of regional and global climate deterioration, such as at 8.6ka BP and 4.2ka BP.

@@

# Bayesian learning and the dynamics of risk perception in the Neolithic {#sec:4}

@@

>>>>

bibliography: bibliography.bib
bibstyle: spbasic
# bibstyle options spbasic(default), spphys, spmpsci
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
header-includes:
- \usepackage{float}
- \floatplacement{figure}{tb}
---

@@

```{r prior, fig.width = 5, fig.height = 3, fig.cap = "Development of an individual's perceived drought risk with time, assuming a fixed drought risk of 0.5. Beliefs are represented as Beta probabiity distributions, and the increased certainty with age reflects the varying effective sample size of the Beta prior."}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

@@

We can treat this problem elegantly using Bayesian statistics using online Bayesian learning from Bernoulli observations (Bissiri 2010). Put simply, the probability of a drought occuring in any given year is treated as a coin flip (i.e. Bernoulli distribution) with parameter $\theta$ representing the risk of drought. Uncertainty in the value of $\theta$ can be represented with a Beta distribution falling between 0 and 1. We can thus represent the exact information content of each individual agent's subjective experience of droughts using the diffusion of the prior belief (Figure \@ref(fig:prior)). What we see here is how an indiviudal's perception of the risk of drought may start off diffuse at first (here as an uniformative flat prior in which all values of theta are equally likely). As each year of experience is observed, the agent updates its beliefs accordingly, balancing the information in this year's observation with the cumulative weight of past experience.

# Holocene drought risk in the eastern Mediterranean {#sec:2}

Estimates of drought risk in the eastern Mediterranean during the the past 10,000 years were derived from the TraCE-21k paleoclimate simulation available on the National Center for Atmospheric Research. TraCE-21K is a state-of-the-art simulation that uses a coupled atmosphere-ocean general circulation model (GCM) to recreate the transient response of the global climate system to changes in the Earth's orbit and greenhouse gasses over the past 22,000 years from the Last Glacial Maximum to the present [@He2011]. It generates physically consistent spatiotemporal climate dynamics, driven by current best estimates of external climate drivers (e.g. orbit, greenhouse gasses, glacial meltwater flux). The model simulates these dynamics on a six hourly timescale, and model outputs are archived at a monthly resolution.

@@

```{r risk, fig.height = 2, fig.width = 6, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period. The dashed line indicates the level of risk beyond which one would plant barley over wheat to maximize subjective expected yields, given the payoffs defined in Table 1.'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  filter(year != min(year)) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, size = 1) +
  geom_col(width = 38)  +
  scale_fill_distiller(palette = 'PuOr') +
  scale_color_distiller(palette = 'PuOr')+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

The bias-corrected wet season precipitation outputs from TraCE-21k show a relatively stationary distribution over the past 10,000 years, with slightly drier conditions during the early-mid Holocene transition at c.a. 8ka BP (Figure \@ref(fig:risk)). The utility of a diversified wheat-barley crop is immediately apparent; rainfall often fails to meet the 300mm threshold for a productive wheat crop during the entire 10,000 year span, but drops below the threshold for barley only a handful of years.

The risk of crop failure due to drought in any given year varies between 10\% and 20\%, punctuated by one to four century-long events where risk drops as low as 6\% and or exceeds 25\% (Figure \@ref(fig:risk)). Centuries of high drought risk cluster around known periods of regional and global climate deterioration, such as at 8.6ka BP and 4.2ka BP.

@@

# The dynamics of risk perception {#sec:4}

@@

What's important to remember here is that both agents are acting rationally based on their prior subjective beliefs, they only differ in those beliefs. So the conservatism of the older agent and the flexibility of the yonger agent are not wrong per se, they simply reflect differet rational approaches to the world.
